A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2015-11-06 imaging_demo/ms.tex
Update sections 4, 5 and 6 with suggested changes
From 137f5a4 (review):
=====================

(*) Update figure 12 (layouts in parameter space)

In ms.tex,

(*) Update sections 4, 5 and 6 with suggested changes

Before the outputs can be compared, we describe the elimination of a minor 
mathematical difference between the two techniques. The squaring operation under 
MOFF imaging in the image plane introduces antenna auto-correlations around the 
zero spacing in the $uv$-plane which are absent in traditional visibility based 
imaging. In order to facilitate a robust comparison between MOFF and visibility 
based imaging techniques, these auto-correlations are removed from both the MOFF 
algorithm output, which is otherwise not an essential part of the core algorithm. 
We describe below how they are removed.

@@

We investigate the two imaging algorithms for differences from the point of 
view of the quality of their outputs. We begin by comparing the gridded 
cross-correlation weights in the $uv$-plane. In MOFF imaging, weights 
from antenna auto-correlations have been removed as described in 
\S\ref{sec:rm-autocorr}.

Fig.~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the MOFF image and the corresponding 
synthesized beam. The sky positions of the simulated sources are indicated by 
solid black circles. The reconstructed sky image has the simulated sources at the
expected sky positions in either case. Both algorithms result in images and 
synthesized beams that are well matched with each other. Their fluxes are 
modulated by a multiplicative power pattern corresponding to that of a uniform
square aperture.

@@

Fig.~\ref{fig:MOFF-FX-uvwts} shows the correlated weights obtained with MOFF 
imaging (left) and visibility based imaging (right). There are two notable 
differences. The first is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The second 
difference is in the weights outside the zero-spacing void. In some regions they 
are different from each other at the few percent level though the sum of weights 
in these regions is the same. For instance, note the difference in weights at 
$(u,v)\approx (2,-10)$. The reason for these differences is described below.

@@

Here we demonstrate our software using narrow band data from the LWA station 
in New Mexico. This data is in LWA narrow-band transient buffer (TBN) format with 
512 voltage time samples from 255 antennas within roughly a diameter of 100~m. 
The data is centered at a frequency of 74.03~MHz, with a sample rate (equal to 
the bandwidth) of 100~kHz with 512 complex time samples in a A/D writeout 
timescale of 5.12~ms, a frequency resolution of 195.3125~Hz and dual 
polarization. There are 391 such time stamps yielding a total duration of 2~s.

@@

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over 20~ms of data and the inner
$\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown in direction 
cosine coordinates -- $l$ along the horizontal axis and $m$ along the vertical 
axis. The flux scale is arbitrary. Even in this proof-of-concept demonstration, 
we see Cyg~A and Cas~A prominently as annotated, thus validating the 
functionality of the EPIC package.

@@

Of all steps in the MOFF pipeline that are repeated for every writeout from 
the F-engine, the slowest step even for dense HERA layouts is found to be the 
spatial two-dimensional FFT in the imaging stage relative to applying the sparse 
matrix gridding convolution, squaring or time-averaging. For instance, even in 
the dense array layout scenario that makes these stages perform the slowest, the 
gridding convolution, squaring and time-averaging take up only $\lesssim 20$\%, 
$\lesssim 20$\% and $\lesssim 5$\% respectively of the total time taken by all 
these processes while the spatial Fourier transform takes up $\gtrsim 55\%$ of
the total time. With sparser arrays the gridding process will be be even faster.

In the visibility based imaging, the predominant computational cost is at
the X-engine requiring $\Nant(\Nant-1)/2$ complex multiplications 
per channel at every correlator writeout timescale.

@@

Fig.~\ref{fig:parameter-space-computations-instruments} shows the number of 
complex operations per frequency channel per cross-polarization per integration 
timescale. Telescopes that fall to the left of the solid line indicate MOFF 
imaging is computationally more efficient than visibility based imaging. All HERA 
layouts except HERA-19 and possibly HERA-37 are in a region of parameter space 
where MOFF imaging holds the advantage. The dashed line showing future trajectory 
of HERA like systems will be clearly favored by MOFF imaging. The dotted line is 
similarly a hypothetical trajectory for the MWA with more tiles added inside the 
same core diameter. The gray shaded area is for a projected LWA expansion and is 
also predominantly in the region favoring MOFF imaging. It is bounded by the LWA1 
and LWA-OV on the left and right respectively. The current (see 
Table~\ref{tab:antenna-layouts}) and a hypothetical expanded layout with a 
four-fold increase in number of elements over a 50\% increase in $b_\textrm{max}$ 
provide the bounds at the bottom and top respectively. Current instruments such 
as MWA and LOFAR lie in parameter space favoring visibility based imaging.

@@

The different antennas are color coded by roughly the class of antenna size 
they fall into. Thus symbols of one color falling into a wedge of the same color 
indicates MOFF imaging is favored for those telescopes and vice versa. For 
e.g., MOFF imaging is favored in HERA-33 and HERA-6769 because they are inside 
the wedge but not so in cases of HERA-19 and HERA-37. A majority of the 
next-generation radio telescopes, namely, HERA-331 and its future expanded 
versions, SKA1-LC, SKA1-LCD, HIRAX, CHIME, LWA1 and its future expansions 
(LWA1-x2x1 and LWA1-x4x1.5 denote two-fold and four-fold increase in the number 
of antennas within a core diameter that is 1 and 1.5 times the current size of 
100~m respectively), and a couple of the expanded versions of LWA-OV (LWA-OVx2x1 
and LWA-OVx4x1.5 denote two-fold and four-fold increase in the number of antennas
within a core diameter that is 1 and 1.5 times the current size of 200~m 
respectively) will fall in the regime where MOFF imaging will be desirable. For 
a fixed baseline length, regions favoring the MOFF algorithm tend to be towards 
large $\Nant$ indicating large-N dense array layouts with smaller antenna 
elements are best suited for deploying EPIC.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding_legended}
  % \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding_legended.eps}
  \caption{Current and planned instruments in parameter space of
    baseline length and number of antennas with MOFF and FX. Line styles of 
    different colors denote different classes of antenna sizes (cyan - 1~m$^2$, 
    blue - 7~m$^2$, purple - 16~m$^2$, green - 28~m$^2$, orange - 150~m$^2$, 
    red - 740~m$^2$, gray - 5900~m$^2$). Solid line style in each color denotes 
    the maximum number of antennas with the corresponding antenna size that can 
    be packed inside various baseline lengths. The region to the right of the 
    solid lines for corresponding antenna size is physically disallowed. Dashed
    lines of each color denote the boundary to the left of which visibility
    based imaging is favored for the corresponding antenna size. Region inside 
    the wedge enclosed by the solid and dashed lines favors using the MOFF 
    algorithm for the corresponding antenna size. These wedges shift leftward 
    with increasing antenna size. The different antennas are color coded by 
    roughly the class of antenna size they fall into. Thus symbols of one 
    color falling into a wedge of the same color indicates MOFF imaging is 
    advantageous for those telescopes and vice versa. For e.g., MOFF imaging
    is favored in HERA-331 and HERA-6769 because they are inside the wedge 
    but not so in cases of HERA-19 and HERA-37.}
  \label{fig:parameter-space-bll-nant-instruments}
\end{figure}

@@

As radio astronomy is entering a new era, advances in instrumentation have to
be accompanied by equal advances in processing techniques to manage 
computational resources. Many future radio telescopes such as the SKA, HERA and 
LWA are headed towards the large-N dense array layout model for which 
computational cost from traditional FX/XF correlator based architecture and 
visibility based imaging starts rising steeply. We have provided the first 
software demonstration of a general purpose imaging algorithm using our generic 
and efficient EPIC software that is designed to bring this cost down from 
$\mathcal{O}(N^2)$ to $\mathcal{O}(N\log_2 N)$. Under the class of direct imaging 
techniques, ours is one of the most generic -- neither does it place any 
constraint on the array layout to be on a regular grid nor does it require the
antenna array to be homogeneous.

Our EPIC package, now publicly available, written in object oriented Python, is 
highly modularized and parallelizable. It includes an implementation of the 
MOFF algorithm in addition to visibility based software holography imaging and 
a data simulator for sky models. It is designed to provide a development 
platform to compare different imaging approaches and serve as a stepping stone 
for real-life GPU/FPGA based implementation on telescopes. It has been 
successfully tested on simulated as well as real LWA observations.

@@

A unique advantage is the instantaneous availability of calibrated time-domain 
images bundled together with the hardware frontend such as the F-engine as an 
integrated module. It is seen to have significant savings in data throughput 
relative to a X-engine based pipeline. Hence, it is a compelling candidate for 
time-domain radio astronomy, e.g. search for and monitoring of transients. 
Transient detection pipeline at the backend of EPIC can be fine-tuned to target 
fast transients such as the Fast Radio Bursts \citep[FRB;][]{tho13} on 
millisecond timescales at GHz frequencies or slow transients from planetary and 
exoplanetary origins at frequencies around 100~MHz.

>>>>

Before the outputs can be compared, we describe the elimination of a minor 
mathematical difference between the two techniques. The squaring operation under 
MOFF imaging in the image plane introduces antenna auto-correlations around the 
zero spacing in the $uv$-plane which are absent in traditional visibility based 
imaging. In order to facilitate a robust comparison between MOFF and visibility 
based imaging techniques, these auto-correlations are removed from the MOFF 
algorithm output, which is otherwise not an essential part of the core algorithm. 
We describe below how they are removed.

@@

We investigate the two imaging algorithms for differences from the point of 
view of the quality of their outputs. We begin by comparing the images produced 
with the two approaches.

Fig.~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the correlated weights and data in the 
$uv$ plane, the MOFF image and the corresponding synthesized beam as 
described in \S\ref{sec:rm-autocorr}. The sky positions of the simulated sources 
are indicated by solid black circles. The reconstructed sky image has the 
simulated sources at the expected sky positions in either case. Both algorithms 
result in images and synthesized beams that are well matched with each other. 
Their fluxes are modulated by a multiplicative power pattern corresponding to 
that of a uniform square aperture.

@@

Now we investigate the gridded cross-correlation weights in the $uv$-plane. 
Fig.~\ref{fig:MOFF-FX-uvwts} shows the correlated weights obtained with MOFF 
imaging (left) and visibility based imaging (right). There are two notable 
differences. The first is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The second 
difference is in the weights outside the zero-spacing void. In some regions they 
are different from each other at the few percent level though the sum of weights 
in these regions is the same. For instance, note the difference in weights at 
$(u,v)\approx (2,-10)$. The reason for these differences is described below.

@@

Here we demonstrate our software using narrow band data from the LWA station 
in New Mexico. This data is in LWA narrow-band transient buffer (TBN) format 
from 255 antennas within roughly a diameter of 100~m. The data is centered at a 
frequency of 74.03~MHz, with a sample rate (equal to the bandwidth) of 100~kHz 
with 512 complex time samples per antenna in a A/D writeout timescale of 
5.12~ms, a frequency resolution of 195.3125~Hz and dual polarization. There are 
391 such writeouts (or time stamps; each contains 512 time samples at 100~kHz 
sampling ) yielding a total duration of 2~s.

@@

Fig.~\ref{fig:LWA-image} shows the image produced with MOFF imaging 
packaged in EPIC after averaging over $\approx$~20~ms (four writeouts) of data 
and the inner $\approx 80$\% of bandwidth (roughly 80~kHz). The image is shown 
in direction cosine coordinates -- $l$ along the horizontal axis and $m$ along 
the vertical axis. The flux scale is arbitrary. Even in this proof-of-concept 
demonstration, we see Cyg~A and Cas~A prominently as annotated, thus validating 
the functionality of the EPIC package.

@@

Of all steps in the MOFF pipeline that are repeated for every writeout from 
the F-engine, the most expensive step even for dense HERA layouts is found to be 
the spatial two-dimensional FFT in the imaging stage relative to applying the 
sparse matrix gridding convolution, squaring or time-averaging. For instance, 
even in the conservative dense array layout scenario that makes these other 
stages even more expensive, the gridding convolution, squaring and time-averaging 
take up only $\lesssim 20$\%, $\lesssim 20$\% and $\lesssim 5$\% respectively of 
the total processing time while the spatial Fourier transform takes up 
$\gtrsim 55\%$ of the total time. With sparser arrays the gridding process will 
be be even faster.

In visibility based imaging, the predominant computational cost is at the 
X-engine requiring $\Nant(\Nant-1)/2$ complex multiplications per channel at 
every A/D writeout timescale.

@@

Fig.~\ref{fig:parameter-space-computations-instruments} shows the number of 
complex operations per frequency channel per cross-polarization per integration 
timescale. Telescopes that fall to the left of the solid line indicate MOFF 
imaging is computationally more efficient than visibility based imaging. All HERA 
layouts except HERA-19 and HERA-37 are in a region of parameter space 
where MOFF imaging holds the advantage. The dashed line showing future trajectory 
of HERA like systems will be clearly favored by MOFF imaging. The dotted line is 
similarly a hypothetical trajectory for the MWA with more tiles added inside the 
same core diameter. The gray shaded area is for a projected LWA expansion and is 
also predominantly in the region favoring MOFF imaging. It is bounded by the LWA1 
and LWA-OV on the left and right respectively. The current (see 
Table~\ref{tab:antenna-layouts}) and a hypothetical expanded layout with a 
four-fold increase in number of elements over a 50\% increase in $b_\textrm{max}$ 
provide the bounds at the bottom and top respectively. Current instruments such 
as MWA and LOFAR lie in parameter space favoring visibility based imaging.

@@

The different antennas are color coded by roughly the class of antenna size 
they fall into. Thus symbols of one color falling into a wedge of the same color 
indicates MOFF imaging is favored for those telescopes and vice versa. For 
e.g., MOFF imaging is favored in HERA-331 and HERA-6769 because they are inside 
the wedge but not so in cases of HERA-19 and HERA-37. A majority of the 
next-generation radio telescopes, namely, HERA-331 and its future expanded 
versions, SKA1-LC, SKA1-LCD, HIRAX, and CHIME will fall in the regime where MOFF 
imaging will be desirable. LWA1 and LWA-OV are already very close to the dividing 
line. Their hypothetical expansions,\footnote{LWA1-x2x1 and LWA1-x4x1.5, and 
LWA-OVx2x1 and LWA-OVx4x1.5 denote two-fold and four-fold increase in the number 
of antennas within a core diameter that is 1 and 1.5 times the current size of 
100~m and 200~m respectively for LWA1 and LWA-OV.} will be in the computational 
regime favored by MOFF. For a fixed baseline length, regions favoring the MOFF 
algorithm tend to be towards large $\Nant$ indicating large-N dense array 
layouts with smaller antenna elements are best suited for deploying EPIC.

\begin{figure}
  \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding_legended}
  % \includegraphics[width=\columnwidth]{MOFF_FX_crossover_baseline_n-antennas_rho_fov_gridding_legended.eps}
  \caption{Current and planned instruments in parameter space of
    baseline length and number of antennas with MOFF and FX. Line styles of 
    different colors denote different classes of antenna sizes (cyan - 1~m$^2$, 
    blue - 7~m$^2$, purple - 16~m$^2$, green - 28~m$^2$, orange - 150~m$^2$, 
    red - 740~m$^2$, gray - 5900~m$^2$). Solid line style in each color denotes 
    the maximum number of antennas with the corresponding antenna size that can 
    be packed inside various baseline lengths. The region to the right of the 
    solid lines for corresponding antenna size is physically disallowed. Dashed
    lines of each color denote the boundary to the left of which visibility
    based imaging is favored for the corresponding antenna size. Region inside 
    the wedge enclosed by the solid and dashed lines favors using the MOFF 
    algorithm for the corresponding antenna size. These wedges shift leftward 
    with increasing antenna size. The different antennas are color coded by 
    roughly the class of antenna size they fall into. Thus symbols of one 
    color falling into a wedge of the same color indicates MOFF imaging is 
    advantageous for those telescopes and vice versa. For example, MOFF imaging
    is favored in HERA-331 and HERA-6769 because they are inside the wedge 
    but not so in cases of HERA-19 and HERA-37.}
  \label{fig:parameter-space-bll-nant-instruments}
\end{figure}

@@

As radio astronomy is entering a new era, advances in instrumentation have to
be accompanied by equal advances in processing techniques to manage 
computational resources. Many future radio telescopes such as the SKA, HERA and 
LWA are headed towards the large-N dense array layout model for which 
computational cost from traditional FX/XF correlator based architecture and 
visibility based imaging starts rising steeply. We have provided the first 
software demonstration of a general purpose imaging algorithm using our generic 
and efficient EPIC software that is designed to bring this cost down from 
$\mathcal{O}(\Nant^2)$ to $\mathcal{O}(\Ngrid\log_2 \Ngrid)$. Under the class of 
direct imaging 
techniques, ours is one of the most generic -- neither does it place any 
constraint on the array layout to be on a regular grid nor does it require the
antenna array to be homogeneous.

Our EPIC package, now publicly available, written in object oriented Python, is 
highly modularized and parallelizable. It includes an implementation of the 
MOFF algorithm in addition to visibility based software holography imaging and 
a data simulator for sky models. It is designed to provide a development 
platform to compare different imaging approaches and serve as a stepping stone 
for real-life GPU/FPGA based implementation on telescopes. It has been 
successfully tested on simulated MWA observations as well as real LWA 
observations.

@@

It is seen to have significant savings in data throughput relative to a 
X-engine based pipeline. A unique advantage is the instantaneous availability 
of calibrated time-domain images at no extra cost. Hence, it is a compelling 
candidate for time-domain radio astronomy, e.g. search for and monitoring of 
transients. Potentially, it could allow on-chip processing thus lowering even 
further the relatively low I/O bandwidth shown in table~\ref{tab:data-rates} 
except when a transient event is detected. Transient detection pipelines at 
the backend of EPIC can be fine-tuned to target fast transients such as the 
Fast Radio Bursts \citep[FRB;][]{tho13} on millisecond timescales at GHz 
frequencies or slow transients from planetary and exoplanetary origins at 
frequencies around 100~MHz.
